Fused regression for multi-source gene regulatory network inference

### 2016-07-29 paper_revisions.tex
slashed and burned intro

Numerous studies have shown that functional conservation exists in gene regulatory networks even across large evolutionary distance \cite{satou2006gene, hinman2009evolution,tanay2005conservation,erwin2009evolution}. Our first method -- fused ridge -- uses an L2 penalty on the differences between \textit{a priori} similar interactions (termed fusion penalty). These constraints favor, but do not require, network configurations in which orthologous genes have similar regulators. Because network inference problems are typically underconstrained, these additional constraints allow data in one species to improve network inference performance in another.

By fusing arbitrary interactions within the networks rather than the networks themselves \cite{}, the method allows pooling of data even when there is not a one-to-one correspondence between genes, which is critical to the method's application to cross-species network inference. This general framework can be extended to an arbitrary number of distinct organisms, each contributing data with only partially overlapping sets of genes. This is an advantage over existing approaches to multi-species network inference, which infer only a sub-network for which orthologs exist in every species \cite{joshi_multi-species_2015}.

This general framework for multi-species network inference has broad applications. Any set of pairwise L2 penalties on the differences between regulatory interactions (equivalent to a multivariate Gaussian prior on networks; see methods) can be optimized by the algorithm we develop. As a result, any biological prior on similarity of regulation can be used in place of orthology, including non one-to-one mappings such the prior on genes within the same operon. We demonstrate improvements in network inference performance with the introduction of penalties based on the expectation that genes in the same operon will have similar regulators.

@@

The main difference between the L2 fusion approach and our new adaptive fusion approach occurs when the difference between presumed analogous interactions is large despite the fusion penalty. This condition means that no network exists which simultaneously provides an adequate explanation of expression data and which conforms to our prior on similarity. In this case, we assume that the constrained interactions have diverged and relax the fusion constraint between them (corresponding to the saturating regime of the penalty).

The resulting cost function is non-convex and difficult to optimize \cite{fan2001variable}; however, we can approximate its solution and obtain deeper insight into functional similarity than is available through strict orthology enforcement or the comparison of separately learned networks.

@@

We test the ability of fused L2 and adaptive fusion to improve network recovery on both synthetic data and on real data. We demonstrate the applications of our method to multi-species network inference, using the bacteria species \texit{Bacillus subtilis} and the distantly related \texti{Bacillus anthracis}; multi-platform network inference using \texit{B. subtilis}; and intra-species fused network inference using operons in \textit{B. subtilis}. We then discuss adaptive fusion and show the technique's ability to identify incorrect orthology information which has been introduced to a biological dataset, suggesting the technique's applicability to discovering neo- and sub-functionalizations.

\section{Results}
We used both synthetic networks and real data to evaluate our methods. For the synthetic data, we generated random pairs of networks in which orthologous genes have similar regulatory interactions, and then sampled gene expression from these networks, which we used to derive learned networks for comparison with the input (true) networks. For real data, we examined recovery of a known gold standard in \textit{Bacillus subtilis} \cite{michna_subtiwikidatabase_2014}.

@@

\subsubsection{Adaptive fusion successfully identifies and unfuses 'neofunctionalized' genes}
Orthology prediction is not alwyas a good proxy for functional conservation \cite{gabaldon_functional_2013, studer_how_2009, nehrt_testing_2011}. 
To allow for orthologous genes to be unfused we implemented an adaptive fusion algorithm that attempts to optimize a nonconvex saturating penalty function on differences between fused interactions (figure \ref{adapt-schematic}). 
Pairs of interactions that are dissimilar even after fusion, which sit in the flat portion of this penalty function, are effectively ``unfused,'' and no further penalty is incurred as differences in interaction weights grow. 
Our network procedure strongly favors similarity of fused interactions, and only ``unfuses'' interactions when their similarity cannot be reconciled with expression data. 
As a result, the ``unfusing'' or relaxation of the fusion penalty on certain constraints is much more direct evidence for neofunctionalization than comparing separately fit networks could provide.

@@

We then compared the recovery of interaction weights which were accurately fused, and recovery of interaction weights which were inaccurately fused due to incorrect orthology information (figure \ref{adaptivefusion}b). 
Because fused L2 heavily penalizes large differences between weights which are predicted to be similar, it is able to retrieve a more accurate network for those interactions with true fusion constraints than by learning networks separately (measured by MSE between the true and inferred interaction weights). 
In this simulation, however, the gains accomplished through fused regression do not extend to those interactions lacking true fusion constraints, and the error remains similar to learning networks separately. When we applied adaptive fusion, we did not observe an improvement in network recovery (relative to fused L2). 
However, we were able to identify fusion constraints reflecting incorrect orthology information that had been provided to the algorithm (Figure \ref{adaptivefusion} subsection). [ZACK - WE NEED TO WRITE MORE ABOUT THIS]

@@

Currently, the most widely used approach to combining datasets for network inference is to learn networks from disparate datasets separately, then rank combine the networks as in Marbach et al \cite{marbach_revealing_2010}.
We included, along with our \textit{B. subtilis} dataset, a previously published dataset containing 269 samples covering 104 conditions, obtained using a different tiling microarray (vs custom microarray) and different strain of \textit{B. subtilis} \cite{nicolas2012condition}. 
We compared performance when learning the networks separately and then rank combining (as in cite Ciofani) to learning the networks simultaneously using fusion regression and show improvement in performance using our fused L2 approach (figure \ref{fusedl2-real}a).

Our approach can also make use of other forms of biological data beyond orthology. Information about the expected similarity of TF-gene interactions can come from knowledge about the promoter region or the structure, for bacteria, of polycistronic transcripts. In bacteria, genes within the same operon are typically under the control of the same promoter \cite{lawrence_shared_2002}. We posited, therefore, that genes within the same operon will be regulated similarly by the same transcription factors. We applied fusion regression by creating fusion constraints between a given transcription factor and genes within the same operon, and showed a boost in \textit{B. subtilis} network recovery using within-species fusion (figure \ref{fusedl2-real}b).

@@

As an example of how this formulation allows pooling of data across multiple sources, consider the case where there is a one-to-one orthology between the species being considered (ie different cell-lines of the same organism). The choice of $\lambda_S$ allows one to interpolate between fitting each network independently ($\lambda_S=0$) and pooling data together as if it came from one source ($\lambda_S=\inf$). In the case of cross-species network inference, there is unlikely to be a complete one-to-one orthology. However, because we create constraints the similarity of individual interactions, rather than on the networks as a whole \cite{parikh2011treegl}, we can pool some information across species even when a small fraction of genes have orthologs.

@@

\textit{B. subtilis} and \textit{B. anthracis} are distantly related bacterial species with limited gene orthology [ZACK - HOW DISTANTLY?]. Nevertheless we show that network recovery in \textit{B. subtilis} can be improved through the inclusion of expression data from \textif{B. anthracis}. 
Many previous methods for cross-species network inference operate on the conserved subset of orthologous genes \cite{dillman_comparative_2015}. 
This assumption may be appropriate with very closely related species, but could not be applied in this domain, where a large fraction (62\% and 67\%) of the \textit{B. subtilis} and \textit{B. anthracis} genomes do not have clear orthologs (and many orthologs have ambiguous many-to-many groupings). 
Our method, in contrast, can obtain improvements in network inference performance even when the conserved subset of genes is small. [DISCUSS RELATIONSHIP TO OTHER PAPERS HERE]

This approach is particularly interesting in light of the diversity of important model organisms used in modern biology. Different model systems provide different advantages and disadvantages for experimental design \cite{stolfi_genetic_2012}, but without a principled mechanism for combining data from multiple sources, it is difficult to fully leverage data obtained from even a slightly different model system.

Because the algorithm we developed can accomodate constraints between arbitrary pairs of regulatory interactions, any biological prior representing information about expected regulatory similarity can be represented, even if the prior provides no information about the magnitude or direction of regulation. 
We demonstrate this flexibility through the novel incorporation of operon structure into the gene-regulatory network inference problem. 
In this application, fusion reflects the assumption that genes in the same operons have similar regulators \cite{lawrence_shared_2002}.

Although it is important to take advantage of the similarities of related organisms for generating improved models of gene regulation, it is also critically important to understand how systems differ from one another. 
Our cross-species network inference method is premised on the assumption that orthologous genes have similar regulators. Existing approaches to the genome-wide testing of this assumption learn regulatory networks separately, then compare to identify conservation \cite{aytes_cross-species_2014}. 
Because network inference is typically underconstrained, fitting a network that describes a particular set of experimental observations amounts to sampling a single network from a large set of networks that fit the data equally (or almost equally) as well. As a result, the existence of a difference between corresponding regulatory interactions in a pair of experimentally derived networks is weak evidence that a difference truly does exist. 
Uncoupled global network inference algorithms are a very weak tool for uncovering evolutionary divergence. 
Our method explicitly favors recovering networks for which evolutionarily corresponding interactions are similar. 
As a result, the failure to obtain networks that confirm evolutionary conservation is stronger evidence that conservation does not exist; the next best network that does exhibit conservation must fit the data much worse to have overcome the bias built into the fusion constraints.

>>>>

Numerous studies have shown that functional conservation exists in gene regulatory networks even across large evolutionary distance \cite{satou2006gene, hinman2009evolution,tanay2005conservation,erwin2009evolution}. Our first method -- fused L2 -- takes advantage of this similariy by imposing an L2 penalty on the differences between \texit{a prior} similar interactions (termed fusion penalty). These constraints favor, but do not require, network configurations in which orthologous genes have similar regulators. Because network inference problems are typically underconstrained, these additional constraints allow data in one species to improve network inference performance in another.

%A closely related algorithm -- fused LASSO -- has been applied to gene regulatory network inference in multiple cell lines, or multiple experimental regimes \cite{omranian2016gene, parikh2011treegl, pierson2015sharing}. These approaches have been shown to perform well in fusing data from multiple sources, but because the fusion is between whole networks rather than individual coefficients, they can only be applied when sources share the same set of expressed genes. This all but rules out cross-species applications, and can present difficulty even when data describing a single species sourced from multiple experimental techniques or platforms. Existing approaches to multi-species network inference have inferred only a sub-network of genes with orthologs in every species \cite{joshi_multi-species_2015}, or fit networks separately  \cite{marbach_revealing_2010}. Our approach allows multiple networks to be inferred with biologically motivated constraints on similarity even when the data sources have a relatively small set of overlapping genes.

%This general framework for multi-species network inference has broad applications, extending beyond the case of combining multiple networks. Any set of pairwise L2 penalties on the differences between regulatory interactions (equivalent to a multivariate Gaussian prior on networks; see methods) can be utilized by the algorithm we develop. As a result, any biological prior on similarity of regulation can be used in place of orthology. For example, we demonstrate improvements in network inference performance with the introduction of penalties based on the expectation that genes in the same operon will have similar regulators.

@@

The main difference between the L2 fusion approach and our new adaptive fusion approach occurs when the difference between presumed analogous interactions is large despite the fusion penalty. This condition means that no network exists which simultaneously provides an adequate explanation of expression data and which conforms to our prior on similarity. In this case, we assume that the constrained interactions have diverged and relax the fusion constraint between them (corresponding to the saturating regime of the penalty). The resulting cost function is non-convex and difficult to optimize \cite{fan2001variable}; however, we can approximate its solution and obtain deeper insight into functional similarity than is available through strict orthology enforcement or the comparison of separately learned networks.

@@

\section{Results}
We test the ability of fused L2 and adaptive fusion to improve network recovery on both synthetic data and on real data. We demonstrate the applications of our method to multi-species network inference, using the bacteria species \texit{Bacillus subtilis} and the distantly related \texti{Bacillus anthracis}; multi-platform network inference using \texit{B. subtilis}; and intra-species fused network inference using operons in \textit{B. subtilis}. We then discuss adaptive fusion and show the technique's ability to identify incorrect orthology information which has been introduced to a biological dataset, suggesting the technique's applicability to discovering neo- and sub-functionalizations.

@@

\subsubsection{Adaptive fusion successfully identifies and unfuses 'neofunctionalized' genes}
Although we expect that orthology provides some evidence of regulatory similarity, we know that it is not a perfect proxy for functional conservation \cite{gabaldon_functional_2013, studer_how_2009, nehrt_testing_2011}. Some gene interactions may be non-conserved, and in order to minimize bias it would be desireable to identify these interactions and relax the fusion constraints on them. To allow for orthologous genes to be unfused we implemented an adaptive fusion algorithm that attempts to optimize a nonconvex saturating penalty function on differences between fused interactions (figure \ref{adapt-schematic}). 
Pairs of interactions that are dissimilar even after fusion, which sit in the flat portion of this penalty function, are effectively ``unfused,'' and no further penalty is incurred as differences in interaction weights grow. 
Our network procedure strongly favors similarity of fused interactions, and only ``unfuses'' interactions when their similarity cannot be reconciled with expression data. 
As a result, the ``unfusing'' or relaxation of the fusion penalty on certain constraints is much more direct evidence for neofunctionalization than comparing separately fit networks could provide.

@@

We then compared the interaction weights which were accurately fused, and interaction weights which were inaccurately fused due to incorrect orthology information (figure \ref{adaptivefusion}b). Because fused L2 heavily penalizes large differences between weights which are predicted to be similar, all learned interaction weights with fusion constraints are similar. Those constraints that are unfused by adaptive fusion are allowed to differ; some of the incorrectly fused constraints have large differences in interaction weights.

@@

Currently, the most widely used approach to combining datasets for network inference is to learn networks from disparate datasets separately, then rank combine the networks as in Marbach et al \cite{marbach_revealing_2010}.
We included, along with our \textit{B. subtilis} dataset, a previously published dataset containing 269 samples covering 104 conditions, obtained using a different tiling microarray (vs custom microarray) and different strain of \textit{B. subtilis} \cite{nicolas2012condition}. 
We compared performance when learning the networks separately and then rank combining (as in \cite{ciofani_validated_2012}) to learning the networks simultaneously using fusion regression, and we show improvement in performance using our fused L2 approach (figure \ref{fusedl2-real}a).

Our approach can also make use of forms of biological data other than RNA expression. Information about the expected similarity of TF-gene interactions can come from knowledge about the promoter region or the structure, for bacteria, of polycistronic transcripts. In bacteria, genes within the same operon are typically under the control of the same promoter \cite{lawrence_shared_2002}. We posited, therefore, that genes within the same operon will be regulated similarly by the same transcription factors. We applied fusion regression by creating fusion constraints between a given transcription factor and genes within the same operon. We also used a subset of the gold-standard as priors by relaxing the $\lambda_R$ parameter for these interactions, as in \cite{greenfield_robust_2013}, and evaluated on the remaining set of priors. Incorporation of operon-derived fusion constraints in \textit{B. subtilis} improved network recovery (figure \ref{fusedl2-real}b).

@@

As an example of how this formulation allows pooling of data across multiple sources, consider the case where there is a one-to-one orthology between the species being considered (ie different cell-lines of the same organism). The choice of $\lambda_S$ allows one to interpolate between fitting each network independently ($\lambda_S=0$) and pooling data together as if it came from one source ($\lambda_S=\inf$). In the case of cross-species network inference, there is unlikely to be a complete one-to-one orthology. However, because we create constraints the similarity of individual interactions, rather than on the networks as a whole \cite{parikh2011treegl, omranian2016gene}, we can pool some information across species even when a small fraction of genes have orthologs.

@@

A number of existing approaches have applied fused regression to related problems in network inference. TreeGL applied fused LASSO to the problem of estimating partial correlations of gene expression in a breast cancer dataset consisting of multiple cell types \cite{parikh2011treegl}. They imposed fusion between the coefficients of models learned for different cell types whenever there was an edge between those cell types in a geneology graph describing the cancer cells' development. A similar approach was recently applied to the problem of inferring TF to gene regulatory weights from data sources describing multiple environmental conditions in \textit{E. Coli}, \textit{Mycobacterium tuberculosis}, and \textit{Mus musculus} \cite{omranian2016gene}. In this formulation, networks associated with each data source were fit simultaneously, with an $L1$ constraint on the similarity of (arbitrarily ordered) adjacent networks. In both of these approaches, fusion was restricted to be between corresponding entries of the coefficient matrices. In terms of our formulation in the cross-species case, this is equivalent to the constraints that would be generated by a one-to-one orthology mapping between species, limiting to the (typical) case where orthology is partial or not one-to-one. By optimizing a more general objective function -- in which fusion constraints can be placed on arbitrary pairs of coefficients associated with otherwise unrelated regression tasks -- we can extend this work to the cross-species case.

\textit{B. subtilis} and \textit{B. anthracis} are distantly related bacterial species with limited gene orthology. Nevertheless we show that network recovery in \textit{B. subtilis} can be improved through the inclusion of expression data from \textif{B. anthracis}. Many previous methods for cross-species network inference operate on the conserved subset of orthologous genes \cite{dillman_comparative_2015}. This assumption may be appropriate with very closely related species, but could not be applied in this domain, where a large fraction (62\% and 67\%) of the \textit{B. subtilis} and \textit{B. anthracis} genomes do not have clear orthologs. Our method, in contrast, can obtain improvements in network inference performance even when the conserved subset of genes is small.

This approach is particularly interesting in light of the diversity of important model organisms used in modern biology. Different model systems provide different advantages and disadvantages for experimental design \cite{stolfi_genetic_2012}, but in many cases work in less used systems is hampered by lack of available data. Our intent is to provide a principled method for combining data from diverse sources, so that results in specialized systems can be integrated with data from well studied organism.

Although it is important to take advantage of the similarities of related organisms for generating improved models of gene regulation, it is also critically important to understand how systems differ from one another. Existing approaches to the genome-wide testing of the assumption that orthologous genes have similar regulators learn regulatory networks separately, then compare to identify conservation \cite{aytes_cross-species_2014}. 
Because network inference is typically underconstrained, fitting a network that describes a particular set of experimental observations amounts to sampling a single network from a large set of networks that fit the data equally (or almost equally) as well. As a result, the existence of a difference between corresponding regulatory interactions in a pair of experimentally derived networks is weak evidence that a difference truly does exist. 
Uncoupled global network inference algorithms are a very weak tool for uncovering evolutionary divergence. 
Our method explicitly favors recovering networks for which evolutionarily corresponding interactions are similar. 
As a result, the failure to obtain networks that confirm evolutionary conservation is much more direct evidence that the networks have truly diverged.
